---
title: "Practical: Introduction to BiSSE and HiSSE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(12)
```

## Source

You can find the source files for this tutorial on the class GitHub repository:

[https://github.com/EEOB-Macroevolution/EEOB565X-Spring2020/tree/master/practicals/03-BiSSE-HiSSE](https://github.com/EEOB-Macroevolution/EEOB565X-Spring2020/tree/master/practicals/03-BiSSE-HiSSE)


## Setup

The BiSSE model is implemented as part of the `diversitree` package, which also contains many other -SSE models, while the HiSSE model is implemented in the standalone package `hisse`. So we will start by installing and loading both of these packages.

```{r eval=FALSE}
install.packages("diversitree")
install.packages("hisse")
```
```{r}
library(diversitree)
library(hisse)
```

## Detecting changes in diversification rates with BiSSE

### Simulating trees

To test the ability of BiSSE to detect differences in diversification rates, we are going to simulate a tree under a character-driven birth-death process, where the birth rates are very dependent on a trait while the death rates are identical. The following will create a tree with 50 tips, with a birth rate of 1.0 in state 0 and 4.0 in state 1, a death rate in both states of 0.5 and transition rates of 0.2 in the direction 0 -> 1 and 0.8 in the direction 1 -> 0.

```{r}
sim_parameters = c(1.0, 4.0, 0.5, 0.5, 0.2, 0.8)
tree_bisse = tree.bisse(sim_parameters, max.taxa = 50)
plot(tree_bisse)
```

We can also show the history of our simulated tree. Here edges in state 0 will be shown in red, and edges in state 1 in blue.

```{r}
treehistory = history.from.sim.discrete(tree_bisse, 0:1)
plot(treehistory, tree_bisse, cols = c("red", "blue"))
```

### Fitting the BiSSE model

Now we will find out if BiSSE is able to correctly infer whether there are changes in the diversification rates on our simulated tree, by estimating the rates using maximum likelihood.

This simply attaches a BiSSE model to our simulated tree:

```{r}
bisse_model = make.bisse(tree_bisse, tree_bisse$tip.state)
```

We are also going to need a starting point for our search:

```{r}
start_mle = starting.point.bisse(tree_bisse)
start_mle
```

Finally, we are ready to perform ML estimation:

```{r}
bisse_mle = find.mle(bisse_model, start_mle)
```

We can look at the final likelihood of the model:

```{r}
logLik(bisse_mle)
```

and compare the ML estimates of the parameters to our simulation parameters:

```{r}
sim_parameters
round(coef(bisse_mle),digits = 2)
```

We can also compare the fit of the BiSSE model to a constrained model where the birth and death rates are identical between states

```{r}
null_model = constrain(bisse_model, lambda0 ~ lambda1)
null_model = constrain(null_model, mu0 ~ mu1)
```

Similarly to above, we perform ML estimation on this model

```{r}
start_mle_null = start_mle[c(-1,-3)]
start_mle_null

null_mle = find.mle(null_model, start_mle_null)
round(coef(null_mle),digits = 2)
```

Let us compare the fit of both models

```{r}
anova(bisse_mle, constrained = null_mle)
```


## Detecting changes in diversification rates with HiSSE

One issue with the BiSSE model is that it has trouble distinguishing between changes in rates that are tied to the character state tested, and changes in rates which are independent of that character.
To test this, we will make up an artificial character for our tree

```{r}
incorrect_tip_states = c(rep(1,25), rep(0,25))
names(incorrect_tip_states) = names(tree_bisse$tip.state)
incorrect_tip_states
```

and run ML estimation using the BiSSE model and this new character

```{r}
incorrect_bisse_model = make.bisse(tree_bisse, incorrect_tip_states)
incorrect_bisse_mle = find.mle(incorrect_bisse_model, start_mle)
```

Similarly to before, we will also try a null model where both birth rates and death rates are identical between states

```{r}
incorrect_null_model = constrain(incorrect_bisse_model, lambda0 ~ lambda1)
incorrect_null_model = constrain(incorrect_null_model, mu0 ~ mu1)
incorrect_null_mle = find.mle(incorrect_null_model, start_mle_null)
```

Finally, we can again compare the fit of both models

```{r}
anova(incorrect_bisse_mle, constrained = incorrect_null_mle)
```

The HiSSE model was developed to address this issue.